---
title: "Two-Acid Bayesian Age Inversion (Asp + Glu)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Two-Acid Bayesian Age Inversion (Asp + Glu)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width  = 7,
  fig.height = 5
)
library(AARP)
library(ggplot2)
library(dplyr)
```

## Overview

The vignettes so far treat AAR as a **paleothermometer**: kinetics are
calibrated from lab heating experiments and a known-temperature core
(`vignette("kinetics-fitting")`), then an unknown temperature history is
inferred from a single amino acid's downcore D/L profile
(`vignette("temperature-reconstruction")`).

This vignette turns the same machinery around and uses it for
**AAR geochronology**: dating a fossil sample from its D/L ratio, given an
independently-dated calibration set. The approach follows an in-prep
calibration study of *Neogloboquadrina pachyderma* (Np), the dominant
planktonic foraminifer in polar and subpolar oceans, which is dated using
aspartic acid (Asp) and glutamic acid (Glu) simultaneously (Kaufman et al.,
in prep.). Bottom water at these high-latitude sites stays close to freezing
across glacial-interglacial cycles, so temperature is treated as
approximately constant and age becomes the unknown to solve for -- the
mirror image of the temperature-reconstruction problem.

Combining two amino acids is valuable because each has its own calibration
scatter, and a sample's true age is more tightly constrained by two partly
independent lines of evidence than by either alone. The original study
combines Asp and Glu ages by **generalized least squares (GLS)**, weighting
each acid's age estimate by its inverse variance and their covariance. That
approach works, but the raw GLS weight can go negative when the two acids
are highly correlated, requiring an ad hoc "softplus" floor to keep both
weights positive.

Here we build a **two-stage Bayesian MCMC model** that reuses this package's
existing `run_mcmc()` sampler and plotting utilities:

- **Stage 1** fits Bayesian TDK (Asp) and SPK (Glu) calibration curves
  jointly, with a correlated bivariate-normal residual structure, to a real
  128-sample Np calibration dataset.
- **Stage 2** infers a full posterior distribution for a new sample's age
  from its Asp and Glu D/L, using the two acids' evidence jointly through
  that same correlated likelihood.

Because this is an ordinary Bayesian posterior rather than an algebraic GLS
combination, the two acids combine coherently for any value of the
correlation -- no floor or smoothing constant is needed -- and the model can
just as easily incorporate a genuine prior (e.g. a stratigraphic age range)
if one is available.

## Calibration models

Following Allen et al. (2013), Asp D/L-age data are best described by
**time-dependent kinetics (TDK)**, and Glu by **simple power-law kinetics
(SPK)**:

$$t_\mathrm{Asp} = a_\mathrm{Asp} \cdot \mathrm{atanh}(D/L_\mathrm{Asp})^{e_\mathrm{Asp}}
\qquad
t_\mathrm{Glu} = a_\mathrm{Glu} \cdot (D/L_\mathrm{Glu})^{e_\mathrm{Glu}}$$

`predict_age_asp()` / `predict_age_glu()` implement these forward
(D/L $\to$ age) transforms; `predict_DL_asp()` / `predict_DL_glu()` are
their inverses.

## The calibration dataset

The package includes the 128-sample Np calibration dataset (Arctic Ocean,
Nordic Seas, North Atlantic, and Southern Ocean cores; ages 2 ka to 1.4 Myr),
compiled for the in-prep calibration study.

```{r load_data}
np_path <- system.file("extdata", "np_calibration.csv", package = "AARP")
np      <- read.csv(np_path)

cal_data <- data.frame(age_ka = np$age_ka, DL_Asp = np$DL_Asp, DL_Glu = np$DL_Glu)
nrow(cal_data)
head(cal_data)
```

```{r plot_data, fig.height=4}
ggplot(np, aes(x = DL_Asp, y = DL_Glu, colour = age_ka)) +
  geom_point(size = 1.8) +
  scale_colour_viridis_c(trans = "log10", name = "Age (ka)") +
  labs(x = "D/L Asp", y = "D/L Glu",
       title = "Np calibration dataset: Asp vs. Glu D/L") +
  theme_bw()
```

## Stage 1: Bayesian calibration fit

Both acids are measured on the same dated samples, so their residuals from
the mean age-D/L trend are expected to covary: a sample that racemizes a bit
faster than average tends to do so in both amino acids. `log_lik_age_calibration()`
models this directly with a bivariate normal likelihood on the two acids'
log-age residuals at each calibration sample, sharing a single correlation
parameter `rho`.

Priors on the six calibration parameters plus `rho` are weakly informative,
centred on typical literature values but wide enough that ~128 samples
dominate the posterior:

```{r prior_table, echo=FALSE}
knitr::kable(data.frame(
  Parameter = c("log(a_Asp)", "e_Asp", "log(sigma_Asp)",
                "log(a_Glu)", "e_Glu", "log(sigma_Glu)", "z_rho"),
  Prior     = c("Normal(log(3000), 3)", "Normal(2.5, 2)", "Normal(log(0.4), 1.5)",
                "Normal(log(9000), 3)", "Normal(2.2, 2)", "Normal(log(0.4), 1.5)",
                "Normal(0, 1)")
))
```

```{r mcmc_stage1, cache=TRUE}
init_cal <- c(log_a_Asp = log(3000), e_Asp = 2.5, log_sigma_Asp = log(0.4),
             log_a_Glu = log(9000), e_Glu = 2.2, log_sigma_Glu = log(0.4),
             z_rho = 0)

prop_cal <- c(log_a_Asp = 0.02, e_Asp = 0.015, log_sigma_Asp = 0.03,
             log_a_Glu = 0.02, e_Glu = 0.015, log_sigma_Glu = 0.03,
             z_rho = 0.06)

set.seed(42)
mcmc_cal <- run_mcmc(log_posterior_age_calibration,
                     init        = init_cal,
                     n_iter      = 40000,
                     proposal_sd = prop_cal,
                     calibration_data = cal_data)

cat("Acceptance rate:", round(mcmc_cal$acceptance_rate * 100, 1), "%\n")
```

### Trace plots

```{r traces_stage1, fig.height=8}
plot_mcmc_chains(mcmc_cal, burnin = 10000)
```

### Posterior calibration constants

```{r summarize_stage1}
calib_params <- summarize_calibration(mcmc_cal, burnin = 10000)
calib_params
```

For comparison, the study's independently fitted OLS constants are
$a_\mathrm{Asp} = 2872$, $e_\mathrm{Asp} = 2.76$,
$a_\mathrm{Glu} = 9581$, $e_\mathrm{Glu} = 2.20$, $\rho = 0.825$. The joint
Bayesian fit recovers a similar correlation and comparable curve shapes; the
scale/exponent pair differs somewhat because the joint likelihood optimizes
both curves' fit *and* their shared residual structure simultaneously,
rather than fitting each acid's OLS regression independently.

### Fitted curves vs. data

```{r plot_fit, fig.height=4.5}
plot_age_calibration_fit(mcmc_cal, cal_data, burnin = 10000)
```

## Stage 2: two-acid age inversion

With the calibration posterior summarised, we can now infer a full age
posterior for any new sample given its Asp and Glu D/L. `invert_two_acid_age()`
runs a one-parameter MCMC chain on $\log(\mathrm{age})$, using
`log_lik_two_acid_age()` to jointly evaluate both acids' evidence through
the shared correlated residual structure fit in Stage 1, and a weak bounded
prior (`log_prior_age()`) that only rules out ages far outside the
calibration range -- easy to tighten if independent age information (e.g.
stratigraphic order) is available.

### Validation against the published GLS calculator

Two worked examples from the study's spreadsheet calculator provide a direct
check: their GLS-combined ages and 66% ("likely") prediction intervals can
be compared against our Bayesian posterior for the same D/L inputs.

```{r stage2_examples, cache=TRUE}
examples <- list(
  "Example 1" = list(DL_Asp = 0.350, DL_Glu = 0.153,
                     gls_median = 176.25, gls_lo66 = 119.47, gls_hi66 = 260.03),
  "Example 2" = list(DL_Asp = 0.272, DL_Glu = 0.110,
                     gls_median = 82.59, gls_lo66 = 49.97, gls_hi66 = 136.52)
)

set.seed(1)
mcmc_examples <- lapply(examples, function(ex) {
  invert_two_acid_age(DL_Asp_obs = ex$DL_Asp, DL_Glu_obs = ex$DL_Glu,
                      calib_params = calib_params, n_iter = 30000)
})

comparison <- do.call(rbind, lapply(names(examples), function(nm) {
  post <- summarize_age_posterior(mcmc_examples[[nm]], burnin = 5000)
  data.frame(sample        = nm,
             bayes_median  = post$median_ka,
             bayes_lo66    = post$lo66_ka,
             bayes_hi66    = post$hi66_ka,
             gls_median    = examples[[nm]]$gls_median,
             gls_lo66      = examples[[nm]]$gls_lo66,
             gls_hi66      = examples[[nm]]$gls_hi66)
}))

knitr::kable(comparison, digits = 1)
```

The Bayesian posterior medians and 66% credible intervals closely track the
spreadsheet's GLS point estimates and prediction intervals, confirming the
joint likelihood is doing the same job as the GLS combination -- while
avoiding its softplus weight-flooring step entirely.

```{r plot_examples, fig.height=4}
plot_age_posterior(mcmc_examples[["Example 1"]], burnin = 5000,
                   compare_ages_ka = c("GLS combined" = 176.25)) +
  labs(subtitle = "Example 1: DL_Asp = 0.350, DL_Glu = 0.153")
```

### Recovering known ages from the calibration set

As a further check, we invert the age of three calibration samples near the
10th, 50th, and 90th percentiles of the dataset's age distribution, using
their measured D/L but *not* their published age, and compare the recovered
posterior to the known value. (The very oldest samples are avoided here: as
in the original study's own leave-one-out cross-validation, the calibration
is least precise -- and mildly biased young -- at its oldest extreme, where
calibration samples are sparsest.)

```{r stage2_known, cache=TRUE}
age_quantiles <- stats::quantile(np$age_ka, c(0.1, 0.5, 0.9))
check_idx     <- sapply(age_quantiles, function(q) which.min(abs(np$age_ka - q)))

set.seed(2)
mcmc_known <- lapply(check_idx, function(i) {
  invert_two_acid_age(DL_Asp_obs = np$DL_Asp[i], DL_Glu_obs = np$DL_Glu[i],
                      calib_params = calib_params, n_iter = 30000)
})

known_summary <- do.call(rbind, lapply(seq_along(check_idx), function(k) {
  i    <- check_idx[k]
  post <- summarize_age_posterior(mcmc_known[[k]], burnin = 5000)
  data.frame(UAL = np$UAL[i], published_age_ka = np$age_ka[i],
             posterior_median_ka = post$median_ka,
             lo66_ka = post$lo66_ka, hi66_ka = post$hi66_ka)
}))

knitr::kable(known_summary, digits = 1)
```

```{r plot_known, fig.height=4}
plot_age_posterior(mcmc_known[[3]], burnin = 5000,
                   true_age_ka = np$age_ka[check_idx[3]]) +
  labs(subtitle = paste0("~90th-percentile-age calibration sample, published age = ",
                         round(np$age_ka[check_idx[3]]), " ka"))
```

The recovered posteriors bracket the published ages within their credible
intervals, as expected for samples that were themselves part of the
calibration fit.

## Why Bayesian?

The GLS approach implemented in the spreadsheet calculator is fast and
transparent, and this vignette confirms our Bayesian model reproduces it
closely. The Bayesian formulation offers a few things the GLS calculator
does not, at the cost of running an MCMC chain instead of evaluating a
formula:

- **No ad hoc weight flooring.** When Asp and Glu are highly correlated, the
  GLS weight on the less-precise acid can go negative, requiring a smoothed
  floor (`s`, `w` in the spreadsheet) to keep the combination well-behaved.
  The joint likelihood here never produces a negative "weight" because there
  isn't one -- both acids simply contribute evidence to one posterior.
- **Priors.** `log_prior_age()` is an ordinary Bayesian prior. If a
  stratigraphic constraint, a minimum/maximum age, or another proxy's age
  estimate is available for a sample, it can be folded in directly by
  tightening `range_ka` or swapping in an informative density, sharpening
  the posterior beyond what the D/L data alone provide.
- **Full posterior, not just symmetric log-normal bands.** The three
  "more likely than not / likely / very likely" IPCC-style bands in the
  spreadsheet are read off an assumed log-normal shape; here they are
  genuine empirical quantiles of the sampled posterior, which can depart
  from log-normality when the prior's age bound becomes informative.
- **Reuses the same MCMC infrastructure** as the temperature-reconstruction
  workflow (`run_mcmc()`, `plot_mcmc_chains()`), so the two problems --
  inferring temperature from a single acid, and inferring age from two --
  are solved with one consistent Bayesian toolkit.

One simplification relative to the full study: the likelihood here uses only
the calibration-scale residual variance (`sigma_Asp`, `sigma_Glu`), not the
additional analytical-measurement-uncertainty term the study adds via each
sample's replicate SE. The study finds that term contributes roughly 1% of
total variance for typical replicate counts, so this simplification has
little practical effect, but a per-sample measurement-uncertainty term could
be added to `log_lik_two_acid_age()` if replicate SEs are known to be
unusually large.
