---
title: "Two-Acid Age Dating: A Shared Forward Model (Asp + Glu)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Two-Acid Age Dating: A Shared Forward Model (Asp + Glu)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width  = 7,
  fig.height = 5
)
library(AARP)
library(ggplot2)
library(dplyr)
```

## Overview

`vignette("two-acid-age-inversion")` builds a Bayesian two-acid age model
that reproduces an in-prep Np AAR age calculator: aspartic acid (Asp) is
fit with a "time-dependent kinetics" (TDK) curve, glutamic acid (Glu) with
a "simple power-law kinetics" (SPK) curve -- two different empirical curve
*forms*, each fit by regressing `log(age)` on `log(f(D/L))`.

This vignette takes a different starting point: **Asp and Glu should
racemize according to the same underlying chemistry** -- the TDK/SPK split
is a curve-fitting convenience for getting the best empirical fit to each
acid, not evidence of different reaction mechanisms. So instead of two
curve forms, this vignette fits **one shared kinetic model** (the same
Arrhenius + Bada power-law model already used for isoleucine elsewhere in
this package: `arrhenius()`, `dRacPL()`, `racemize_one_depth()`) to each
acid independently, each getting its own `Ae` (pre-exponential factor),
`Ea` (activation energy), and `x` (power-law exponent), but evaluated
through the *same* forward-model code path.

Rather than a single-shot "effective temperature" calculation, D/L is
forward-simulated through the package's actual depth-series machinery:
a real per-core age-depth model (`make_age_model()`/`rate_to_age_model()`)
and a temperature *history* (not a scalar), using `racemize_one_depth()`
exactly as it's used for lake-sediment paleothermometry.

**What this buys us over the curve fit:** an activation energy in real
physical units (kcal/mol), directly comparable across acids and to the
isoleucine kinetics fit in `vignette("kinetics-fitting")`, and a model that
in principle can use whatever temperature history is actually known for a
core, rather than assuming one fixed effective value forever.

**What we don't have (yet), and simulate as an explicit placeholder:** the
calibration dataset records only one representative bottom-water
temperature per sample, not an actual paleotemperature reconstruction, and
most cores have too few samples to build a well-resolved age-depth model.
Both gaps are filled here with clearly-labeled placeholders -- a synthetic
mild glacial-interglacial temperature oscillation, and a straight-line
age-depth model for single-sample cores -- meant to be swapped for real
inputs as they become available.

## Parameter glossary

Model parameters accumulate quickly across the two stages below. This table
collects all of them in one place for reference.

```{r param_glossary, echo=FALSE}
knitr::kable(data.frame(
  Parameter = c(
    "`Ae`", "`Ea`", "`x`", "`sigma`", "`rho`",
    "`log_age`", "`temp_offset`", "`depth_cm`", "`dT_yr`"
  ),
  Stage = c(
    "Stage 1", "Stage 1", "Stage 1", "Stage 1", "Stage 1",
    "Stage 2", "Stage 2 (temp. uncertainty)", "Stage 2", "Both"
  ),
  Meaning = c(
    "Arrhenius pre-exponential (frequency) factor, one value per acid. Sets the overall reaction rate scale.",
    "Arrhenius activation energy (kcal/mol), one value per acid. Controls how strongly the rate depends on temperature.",
    "Bada power-law exponent, one value per acid. Same role as `x` in `default_AAR_params` for isoleucine: linearizes D/L accumulation as `(D/L)^x` (or `atanh` for the empirical TDK model in the companion vignette).",
    "Residual SD of observed vs. forward-model-predicted D/L, one value per acid (D/L units, not log-age units).",
    "Correlation between the two acids' D/L residuals at a given calibration sample (both acids racemize under the same history, so their departures from the model trend co-vary).",
    "The single Stage 2 unknown: natural log of the candidate sample age (ka BP). Priors and posteriors are reported back-transformed to ka.",
    "Nuisance parameter (deg C) added to a sample's recorded `temp_C` to represent uncertainty in its true effective bottom-water temperature; only used in the temperature-uncertainty sensitivity analysis below.",
    "Depth of a sample below the sediment surface. For Stage 1 (calibration) this is a real recorded depth; for Stage 2 (a new sample of unknown age) it is an arbitrary placeholder that cancels out of the single-point age-depth model.",
    "Integration timestep (yr) used by `racemize_one_depth()` to numerically accumulate D/L over the age-depth model and simulated temperature history."
  )
))
```

## Age-depth models and simulated temperature histories

```{r load_data}
np_path <- system.file("extdata", "np_calibration.csv", package = "AARP")
np      <- read.csv(np_path)
# A handful of rows record "-" for SD (single-subsample entries with no
# replicate spread); read.csv leaves the whole column as character as a
# result, so coerce explicitly.
np$SD_Asp <- suppressWarnings(as.numeric(np$SD_Asp))
np$SD_Glu <- suppressWarnings(as.numeric(np$SD_Glu))
nrow(np)
```

`build_core_age_models()` groups samples by core and builds a real
interpolated age-depth model wherever a core has two or more dated levels
(e.g. study `"Kaufman et al., 2013"`, core `"Link 16"`, with four samples
from 0.1 to 3.9 mbsf); single-sample cores fall back to a straight line
from the core top through that one point.

```{r age_depth_example, fig.height=3.5}
pc <- precompute_forward_inputs(np)

# A multi-sample core
core_key <- paste(np$study[!is.na(np$temp_C)], np$core[!is.na(np$temp_C)])
multi_i  <- which(core_key == names(sort(-table(core_key)))[1])

depth_seq <- seq(0, max(pc$depth_cm[multi_i]), length.out = 100)
age_seq   <- pc$age_model[[multi_i[1]]]$depth_to_age(depth_seq)

ggplot(data.frame(depth_cm = depth_seq, age_ka = age_seq),
       aes(x = depth_cm, y = age_ka)) +
  geom_line(colour = "steelblue") +
  geom_point(data = data.frame(depth_cm = pc$depth_cm[multi_i],
                               age_ka = np$age_ka[!is.na(np$temp_C)][multi_i]),
             size = 2) +
  labs(x = "Depth (cm)", y = "Age (ka BP)",
       title = "Example per-core age-depth model") +
  theme_bw()
```

`simulate_temp_history()` generates the placeholder temperature history: a
mild (~1&deg;C amplitude, ~100 ka period) oscillation around each sample's
recorded temperature, giving the depth-series time integration a genuine
history to integrate over instead of a constant.

```{r temp_history_example, fig.height=3.5}
th <- simulate_temp_history(temp_C = -0.5, max_age_ka = 400)

ggplot(th, aes(x = age_ka, y = temp_C)) +
  geom_line(colour = "darkorange") +
  scale_x_reverse() +
  labs(x = "Age (ka BP)", y = "Temperature (°C)",
       title = "Simulated placeholder temperature history",
       subtitle = "Anchored on a recorded temp_C of -0.5°C -- not a real reconstruction") +
  theme_bw()
```

## Stage 1: forward-model calibration fit

`precompute_forward_inputs()` builds the (depth, age-model,
temperature-history) triple for every calibration sample **once**, since
only the kinetic parameters (`Ae`, `Ea`, `x` per acid) change between MCMC
iterations -- the age models and temperature histories don't depend on
them.

```{r prior_table, echo=FALSE}
knitr::kable(data.frame(
  Parameter = c("log(Ae_Asp), log(Ae_Glu)", "Ea_Asp, Ea_Glu", "x_Asp, x_Glu",
                "log(sigma_Asp), log(sigma_Glu)", "z_rho"),
  Prior     = c("Normal(42, 5)", "Normal(30, 3) -- informed", "Normal(3, 1)",
                "Normal(log(0.03), 1.5)", "Normal(0, 1)")
))
```

`Ea` uses a deliberately **informed** prior: the calibration dataset's
natural bottom-water temperatures span only about 3.5&deg;C, far too narrow
to identify an activation energy from field data alone (the same limitation
the original study notes for the Arrhenius/heating-experiment comparison
approach). The prior is centred in the ~25-35 kcal/mol range typical of
published racemization activation energies for calcite-hosted amino acids.

```{r mcmc_stage1, cache=TRUE}
init_fwd <- c(log_Ae_Asp = 42, Ea_Asp = 30, x_Asp = 3, log_sigma_Asp = log(0.03),
             log_Ae_Glu = 42, Ea_Glu = 30, x_Glu = 3, log_sigma_Glu = log(0.03),
             z_rho = 0)

prop_fwd <- c(log_Ae_Asp = 0.02, Ea_Asp = 0.015, x_Asp = 0.008, log_sigma_Asp = 0.02,
             log_Ae_Glu = 0.02, Ea_Glu = 0.015, x_Glu = 0.008, log_sigma_Glu = 0.02,
             z_rho = 0.03)

set.seed(42)
mcmc_fwd <- run_mcmc(log_posterior_age_calibration_fwd,
                     init        = init_fwd,
                     n_iter      = 10000,
                     proposal_sd = prop_fwd,
                     precomputed = pc,
                     dT_yr       = 2000)

cat("Acceptance rate:", round(mcmc_fwd$acceptance_rate * 100, 1), "%\n")
```

### Trace plots

```{r traces_stage1, fig.height=9}
plot_mcmc_chains(mcmc_fwd, burnin = 2000)
```

### Posterior calibration constants

```{r summarize_stage1}
calib_params_fwd <- summarize_calibration_fwd(mcmc_fwd, burnin = 2000)
calib_params_fwd
```

### Prior vs. posterior for every parameter

```{r plot_priorpost, fig.height=8}
plot_prior_posterior_fwd(mcmc_fwd, burnin = 2000)
```

`x_Asp`, `x_Glu`, and both `log_sigma` parameters narrow substantially
relative to their priors -- the data are informative about the shape of the
D/L-age relationship and the residual scatter. `Ea_Asp` and `Ea_Glu`, by
contrast, barely move from their prior: exactly the identifiability
limitation the compensation-ridge plot below shows from a different angle.

### Fitted vs. observed D/L

```{r plot_fit, fig.height=4}
plot_age_calibration_fit_fwd(mcmc_fwd, pc, burnin = 2000)
```

### The Ea identifiability problem, made visible

```{r plot_ridge, fig.height=4}
plot_ae_ea_posterior(mcmc_fwd, burnin = 2000)
```

`Ae` and `Ea` trade off almost perfectly along a ridge -- the classic
Arrhenius "compensation effect." With only ~3.5&deg;C of natural temperature
range in this dataset, many `(Ae, Ea)` pairs predict nearly the same
racemization rate, so the posterior mean `Ea` above sits close to its prior
centre (30 kcal/mol) rather than being pinned down by the data. This is an
honest finding, not a fitting failure: it's exactly the same limitation the
original study flags for its own Arrhenius comparison, and it's the reason
the AAGL heating-experiment work described in `vignette("kinetics-fitting")`
matters for this problem too -- controlled multi-temperature experiments are
what actually resolve `Ae`/`Ea` separately.

## Stage 2: forward-model age inversion

```{r stage2_examples, cache=TRUE}
mean_temp <- mean(np$temp_C, na.rm = TRUE)

examples <- list(
  "Example 1" = list(DL_Asp = 0.350, DL_Glu = 0.153),
  "Example 2" = list(DL_Asp = 0.272, DL_Glu = 0.110)
)

set.seed(1)
mcmc_examples_fwd <- lapply(examples, function(ex) {
  invert_two_acid_age_fwd(DL_Asp_obs = ex$DL_Asp, DL_Glu_obs = ex$DL_Glu,
                          calib_params_fwd = calib_params_fwd,
                          temp_C = mean_temp, n_iter = 15000)
})
```

Darrell's spreadsheet examples don't come with a recorded depth or
temperature, so the mean dataset temperature (`r round(mean_temp, 2)`&deg;C)
and an arbitrary placeholder depth are used (the depth cancels out of the
single-point age-depth model; see `?log_lik_two_acid_age_fwd`).

```{r stage2_known, cache=TRUE}
age_quantiles <- stats::quantile(np$age_ka, c(0.1, 0.5, 0.9))
check_idx     <- sapply(age_quantiles, function(q) which.min(abs(np$age_ka - q)))

set.seed(2)
mcmc_known_fwd <- lapply(check_idx, function(i) {
  temp_i <- np$temp_C[i]
  if (is.na(temp_i)) temp_i <- mean_temp
  invert_two_acid_age_fwd(DL_Asp_obs = np$DL_Asp[i], DL_Glu_obs = np$DL_Glu[i],
                          calib_params_fwd = calib_params_fwd, temp_C = temp_i,
                          depth_cm = np$depth_mbsf[i] * 100, n_iter = 15000)
})
```

These are the same three known-age calibration samples (10th/50th/90th
percentile ages) used for validation in `vignette("two-acid-age-inversion")`,
now using each sample's own recorded depth and temperature.

### Comparison: Darrell's GLS calculator vs. both Bayesian approaches

Three independent methods, side by side, for the same five samples: Darrell's
published GLS/softplus point-estimate calculator (`predict_age_gls()`, a
faithful R port of the spreadsheet's own formulas -- see
`?gls_calibration_constants`), the empirical curve-fit Bayesian model from
`vignette("two-acid-age-inversion")`, and the forward model built above.

```{r comparison_table, echo=FALSE}
# GLS reference method (Darrell's spreadsheet calculator, reproduced exactly)
gls_inputs <- data.frame(
  sample = c("Example 1", "Example 2", "10th pct (UAL 7440)",
            "50th pct (UAL 8188)", "90th pct (UAL 8194)"),
  DL_Asp = c(0.350, 0.272, np$DL_Asp[check_idx[1]], np$DL_Asp[check_idx[2]], np$DL_Asp[check_idx[3]]),
  SD_Asp = c(0.010, 0.063, np$SD_Asp[check_idx[1]], np$SD_Asp[check_idx[2]], np$SD_Asp[check_idx[3]]),
  N_Asp  = c(5,     4,     np$n_used[check_idx[1]],  np$n_used[check_idx[2]],  np$n_used[check_idx[3]]),
  DL_Glu = c(0.153, 0.110, np$DL_Glu[check_idx[1]], np$DL_Glu[check_idx[2]], np$DL_Glu[check_idx[3]]),
  SD_Glu = c(0.015, 0.035, np$SD_Glu[check_idx[1]], np$SD_Glu[check_idx[2]], np$SD_Glu[check_idx[3]]),
  N_Glu  = c(5,     4,     np$n_used[check_idx[1]],  np$n_used[check_idx[2]],  np$n_used[check_idx[3]])
)

gls_summary <- do.call(rbind, lapply(seq_len(nrow(gls_inputs)), function(k) {
  gi  <- gls_inputs[k, ]
  res <- predict_age_gls(gi$DL_Asp, gi$SD_Asp, gi$N_Asp, gi$DL_Glu, gi$SD_Glu, gi$N_Glu)
  data.frame(sample = gi$sample, gls_median = res$t_comb,
            gls_lo66 = res$lo66, gls_hi66 = res$hi66)
}))

# Empirical-model reference values (from vignette("two-acid-age-inversion"))
empirical <- data.frame(
  sample       = c("Example 1", "Example 2", "10th pct (UAL 7440)",
                   "50th pct (UAL 8188)", "90th pct (UAL 8194)"),
  published_ka = c(NA, NA, 11.6, 88.8, 631.6),
  emp_median   = c(167.8, 84.3, 12.8, 114.9, 727.3),
  emp_lo66     = c(112.8, 56.9, 8.6, 77.0, 487.2),
  emp_hi66     = c(252.0, 125.4, 19.0, 171.2, 1088.5)
)

fwd_summary <- rbind(
  do.call(rbind, lapply(names(mcmc_examples_fwd), function(nm) {
    s <- summarize_age_posterior(mcmc_examples_fwd[[nm]], burnin = 3000)
    data.frame(sample = nm, fwd_median = s$median_ka,
               fwd_lo66 = s$lo66_ka, fwd_hi66 = s$hi66_ka)
  })),
  do.call(rbind, lapply(seq_along(check_idx), function(k) {
    i <- check_idx[k]
    s <- summarize_age_posterior(mcmc_known_fwd[[k]], burnin = 3000)
    data.frame(sample = paste0(c("10th pct (UAL ", "50th pct (UAL ", "90th pct (UAL ")[k],
                               np$UAL[i], ")"),
               fwd_median = s$median_ka, fwd_lo66 = s$lo66_ka, fwd_hi66 = s$hi66_ka)
  }))
)
fwd_summary$sample <- gls_inputs$sample  # align labels (fwd_summary spells out UAL numbers)

comparison <- merge(merge(gls_summary, empirical, by = "sample", sort = FALSE),
                    fwd_summary, by = "sample", sort = FALSE)
knitr::kable(comparison, digits = 1)
```

```{r plot_comparison, fig.height=4.5, echo=FALSE}
plot_df <- rbind(
  transform(comparison, model = "Darrell's GLS calculator",
           median = gls_median, lo66 = gls_lo66, hi66 = gls_hi66)[
    c("sample", "model", "median", "lo66", "hi66", "published_ka")],
  transform(comparison, model = "Empirical Bayesian (curve fit)",
           median = emp_median, lo66 = emp_lo66, hi66 = emp_hi66)[
    c("sample", "model", "median", "lo66", "hi66", "published_ka")],
  transform(comparison, model = "Forward-model Bayesian (shared kinetics)",
           median = fwd_median, lo66 = fwd_lo66, hi66 = fwd_hi66)[
    c("sample", "model", "median", "lo66", "hi66", "published_ka")]
)

ggplot(plot_df, aes(x = sample, y = median, colour = model)) +
  geom_pointrange(aes(ymin = lo66, ymax = hi66),
                  position = position_dodge(width = 0.5)) +
  geom_point(aes(y = published_ka), colour = "black", shape = 4, size = 2.5,
            na.rm = TRUE) +
  scale_y_log10() +
  labs(x = NULL, y = "Age (ka BP, log scale)",
       title = "Three independent age-dating methods, five samples",
       subtitle = "X = published age (known-age samples only); bars = 66% interval") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 20, hjust = 1),
        legend.position = "bottom")
```

## How much does a second acid help?

Both the empirical curve fit and the forward model combine Asp and Glu
through a *correlated* bivariate likelihood, because both acids are
measured on the same dated sample and racemize under the same history --
their departures from the model trend aren't independent (posterior
`rho` &asymp; `r round(calib_params_fwd$rho, 2)`). High correlation between
two lines of evidence means combining them helps less than it would if they
were independent: this section quantifies that directly by inverting age
from Asp alone, Glu alone, and both acids jointly, for all five samples used
in the method comparison above.

```{r stage2_single_acid, cache=TRUE}
case_studies <- list(
  "Example 1" = list(DL_Asp = 0.350, DL_Glu = 0.153,
                     temp_C = mean_temp, depth_cm = 100,
                     published_ka = NA,
                     both = mcmc_examples_fwd[["Example 1"]]),
  "Example 2" = list(DL_Asp = 0.272, DL_Glu = 0.110,
                     temp_C = mean_temp, depth_cm = 100,
                     published_ka = NA,
                     both = mcmc_examples_fwd[["Example 2"]])
)
for (k in seq_along(check_idx)) {
  i    <- check_idx[k]
  nm   <- gls_inputs$sample[2 + k]  # "10th/50th/90th pct (UAL ...)"
  case_studies[[nm]] <- list(
    DL_Asp   = np$DL_Asp[i], DL_Glu = np$DL_Glu[i],
    temp_C   = ifelse(is.na(np$temp_C[i]), mean_temp, np$temp_C[i]),
    depth_cm = np$depth_mbsf[i] * 100,
    published_ka = np$age_ka[i],
    both     = mcmc_known_fwd[[k]]
  )
}
```

`"Example 1"` and `"Example 2"` are Darrell's spreadsheet demonstration
inputs, not real dated samples -- there's no published age to check them
against. The other three are the same real calibration samples used
throughout this vignette (10th/50th/90th percentile ages), each with a
genuine published age to compare the posteriors against.

```{r stage2_single_acid_run, cache=TRUE}

set.seed(3)
single_acid_results <- lapply(case_studies, function(cs) {
  list(
    Asp  = invert_one_acid_age_fwd(cs$DL_Asp, acid = "Asp",
                                   calib_params_fwd = calib_params_fwd,
                                   temp_C = cs$temp_C, depth_cm = cs$depth_cm,
                                   n_iter = 10000),
    Glu  = invert_one_acid_age_fwd(cs$DL_Glu, acid = "Glu",
                                   calib_params_fwd = calib_params_fwd,
                                   temp_C = cs$temp_C, depth_cm = cs$depth_cm,
                                   n_iter = 10000),
    Both = cs$both
  )
})
```

```{r stage2_single_acid_table, echo=FALSE}
acid_comparison <- do.call(rbind, lapply(names(single_acid_results), function(cs_name) {
  do.call(rbind, lapply(names(single_acid_results[[cs_name]]), function(mn) {
    s <- summarize_age_posterior(single_acid_results[[cs_name]][[mn]], burnin = 2000)
    data.frame(sample = cs_name, acid_used = mn, median_ka = s$median_ka,
               lo66_ka = s$lo66_ka, hi66_ka = s$hi66_ka,
               log_width_66 = log(s$hi66_ka / s$lo66_ka))
  }))
}))
knitr::kable(acid_comparison, digits = c(0, 0, 1, 1, 1, 2))
```

```{r plot_single_acid, fig.height=9, echo=FALSE}
acid_density_df <- do.call(rbind, lapply(names(single_acid_results), function(cs_name) {
  do.call(rbind, lapply(names(single_acid_results[[cs_name]]), function(mn) {
    mc     <- single_acid_results[[cs_name]][[mn]]
    burnin <- if (mn == "Both") 3000 else 2000
    age_ka <- exp(utils::tail(mc$samples[, "log_age"], nrow(mc$samples) - burnin))
    data.frame(sample = cs_name, acid_used = mn, age_ka = age_ka)
  }))
}))
acid_density_df$sample <- factor(acid_density_df$sample, levels = names(case_studies))

published_df <- do.call(rbind, lapply(names(case_studies), function(nm) {
  data.frame(sample = nm, published_ka = case_studies[[nm]]$published_ka)
}))
published_df <- published_df[!is.na(published_df$published_ka), ]

ggplot(acid_density_df, aes(x = age_ka, fill = acid_used)) +
  geom_density(alpha = 0.4, colour = NA) +
  geom_vline(data = published_df, aes(xintercept = published_ka),
            colour = "black", linetype = "dashed", linewidth = 0.8) +
  scale_x_log10() +
  facet_wrap(~sample, scales = "free", ncol = 2) +
  labs(x = "Age (ka BP, log scale)", y = "Posterior density", fill = "Acid(s) used",
       title = "Asp-only vs. Glu-only vs. both acids jointly",
       subtitle = "Dashed line = published age (three known-age samples only; the two spreadsheet Examples have none)") +
  theme_bw()
```

Using both acids narrows the 66% interval relative to the *better* single
acid (usually Asp, whose smaller residual scatter gives it more weight), but
only modestly -- a real gain, not a doubling of precision, because the high
Asp-Glu correlation means Glu mostly confirms what Asp already indicates
rather than adding independent information. Glu alone is consistently the
least precise: its shallower forward-model sensitivity (smaller `x_Glu`)
means a given D/L uncertainty maps to a wider age range. This mirrors the
original study's own finding that Glu typically receives a smaller GLS
weight than Asp.

### Deep dive: Why is the combined estimate sometimes *older* than either acid alone?

Both case studies show the same surprising pattern: the joint (Asp+Glu)
posterior isn't between the two single-acid posteriors, or centred near the
more precise one -- it's older than *both*. That's not a bug; it falls out
of the correlated likelihood, and is worth deriving explicitly for the
UAL 8188 case.

```{r explain_combo, cache=TRUE}
cs <- case_studies[["50th pct (UAL 8188)"]]

params_Asp <- list(Ae = calib_params_fwd$Ae_Asp, Ea = calib_params_fwd$Ea_Asp,
                   R = 0.001987, x = calib_params_fwd$x_Asp)
params_Glu <- list(Ae = calib_params_fwd$Ae_Glu, Ea = calib_params_fwd$Ea_Glu,
                   R = 0.001987, x = calib_params_fwd$x_Glu)

age_grid   <- exp(seq(log(10), log(300), length.out = 200))
resid_grid <- do.call(rbind, lapply(age_grid, function(age_ka) {
  am <- rate_to_age_model(rate_cm_per_ka = cs$depth_cm / age_ka,
                          max_depth_cm   = cs$depth_cm)
  tm <- simulate_temp_history(cs$temp_C, age_ka)
  DL_Asp_pred <- racemize_one_depth(cs$depth_cm, am, tm, dT_yr = 2000,
                                    AAR_params = params_Asp)$DL
  DL_Glu_pred <- racemize_one_depth(cs$depth_cm, am, tm, dT_yr = 2000,
                                    AAR_params = params_Glu)$DL
  data.frame(age_ka = age_ka,
             z_Asp = (cs$DL_Asp - DL_Asp_pred) / calib_params_fwd$sigma_Asp,
             z_Glu = (cs$DL_Glu - DL_Glu_pred) / calib_params_fwd$sigma_Glu)
}))

rho <- calib_params_fwd$rho
resid_grid$Q <- (resid_grid$z_Asp^2 - 2 * rho * resid_grid$z_Asp * resid_grid$z_Glu +
                 resid_grid$z_Glu^2) / (1 - rho^2)

age_Asp_only <- age_grid[which.min(abs(resid_grid$z_Asp))]
age_Glu_only <- age_grid[which.min(abs(resid_grid$z_Glu))]
age_joint    <- age_grid[which.min(resid_grid$Q)]

cat("Asp-only implied age:", round(age_Asp_only), "ka\n")
cat("Glu-only implied age:", round(age_Glu_only), "ka\n")
cat("Joint (min-Q) age:   ", round(age_joint), "ka\n")
```

```{r plot_explain_combo, fig.height=4.5, echo=FALSE}
resid_long <- tidyr::pivot_longer(resid_grid, c("z_Asp", "z_Glu"),
                                  names_to = "acid", values_to = "z")

ggplot(resid_long, aes(x = age_ka, y = z, colour = acid)) +
  geom_hline(yintercept = 0, colour = "grey60") +
  geom_vline(xintercept = age_Asp_only, colour = "#F8766D", linetype = "dashed") +
  geom_vline(xintercept = age_Glu_only, colour = "#00BFC4", linetype = "dashed") +
  geom_vline(xintercept = age_joint, colour = "black", linewidth = 0.8) +
  geom_line(linewidth = 0.9) +
  scale_x_log10() +
  labs(x = "Candidate age (ka BP, log scale)", y = "Standardized residual (z)",
       colour = "Acid",
       title = "Standardized residuals vs. candidate age (UAL 8188)",
       subtitle = "Dashed = each acid's own zero-crossing; solid black = joint (min-Q) age") +
  theme_bw()
```

At any candidate age, each acid contributes a standardized residual
(`z_Asp`, `z_Glu`): how many residual SDs the observed D/L sits from the
model's prediction. A single acid's best-fit age is simply where its `z`
crosses zero. Here those two ages disagree substantially -- Asp alone points
to `r round(age_Asp_only)` ka, Glu alone to `r round(age_Glu_only)` ka --
which is itself informative: the sample is mildly discordant between the two
proxies.

The two-acid likelihood isn't a weighted average of the two single-acid
likelihoods; it's a *correlated* bivariate normal, whose quadratic form is

$$Q = \frac{z_\mathrm{Asp}^2 - 2\rho\, z_\mathrm{Asp} z_\mathrm{Glu} + z_\mathrm{Glu}^2}{1-\rho^2}$$

with `rho` &asymp; `r round(rho, 2)` here. The cross-term `-2*rho*z_Asp*z_Glu`
is the key: it **subtracts** from `Q` (raising the likelihood) whenever
`z_Asp` and `z_Glu` share the same sign, and **adds** to `Q` (lowering the
likelihood) whenever they have opposite signs. Between the two acids' own
optima -- exactly the "compromise" range a simple average would favour --
the residuals necessarily have *opposite* signs (one acid is satisfied, the
other isn't yet), so that entire region is actively penalized by the
positive correlation. The joint posterior is pushed past whichever
single-acid optimum is closer, out to where both residuals flip to
agreeing in sign: here, that's just beyond the Asp optimum, at
`r round(age_joint)` ka.

In plain terms: because Asp and Glu are known to drift together (`rho`
&asymp; 0.88), the model treats "Asp says X, Glu says something else" not
as "split the difference" but as "both are probably being pushed the same
direction by something not in the model, and whichever age makes that
displacement consistent between the two acids is more likely than any age
where they'd have to be wrong in opposite directions." That's a genuine
consequence of modelling the correlation coherently rather than combining
two point estimates by hand -- but it's also a reason to treat `rho` itself
(and the calibration curves it's estimated alongside) with some scepticism:
if the fitted correlation is too high, this mechanism will overreact to
ordinary discordance between two proxies exactly like this.

## How much does the temperature history matter?

Every result so far conditions on one simulated temperature history per
sample. But the calibration dataset only records a single representative
temperature -- the *true* effective temperature could plausibly sit
anywhere in a narrow, cold range around it. `invert_two_acid_age_tempunc()`
folds that uncertainty directly into the age inversion: instead of fixing
the temperature history's anchor, it adds a free `temp_offset` parameter
(prior `Normal(0, 1)` &deg;C) sampled jointly with `log_age`, so the age
posterior automatically marginalizes over plausible temperature histories
rather than conditioning on a single guess.

```{r stage2_tempunc, cache=TRUE}
set.seed(4)
tempunc_result <- invert_two_acid_age_tempunc(
  DL_Asp_obs = 0.350, DL_Glu_obs = 0.153, calib_params_fwd = calib_params_fwd,
  temp_C_center = mean_temp, n_iter = 15000,
  proposal_sd = c(log_age = 0.3, temp_offset = 0.5), temp_sd = 1
)
cat("Acceptance rate:", round(tempunc_result$acceptance_rate * 100, 1), "%\n")

post_tu <- tempunc_result$samples[3001:15000, ]
cat("Posterior temp_offset: mean =", round(mean(post_tu[, "temp_offset"]), 2),
    ", sd =", round(sd(post_tu[, "temp_offset"]), 2), "°C (prior sd = 1°C)\n")
```

```{r plot_tempunc, fig.height=4}
fixed_df <- data.frame(
  age_ka = exp(mcmc_examples_fwd[["Example 1"]]$samples[3001:15000, "log_age"]),
  model  = "Fixed temperature"
)
tempunc_df <- data.frame(
  age_ka = exp(post_tu[, "log_age"]),
  model  = "Temperature uncertainty folded in"
)

ggplot(rbind(fixed_df, tempunc_df), aes(x = age_ka, fill = model)) +
  geom_density(alpha = 0.4) +
  scale_x_log10() +
  labs(x = "Age (ka BP, log scale)", y = "Posterior density",
       title = "Effect of temperature-history uncertainty on the age posterior",
       fill = NULL) +
  theme_bw() +
  theme(legend.position = "bottom")
```

The posterior on `temp_offset` barely moves from its prior (mean near zero,
SD close to the prior SD of 1&deg;C) -- confirming the suspicion that a
single sample's D/L alone cannot meaningfully constrain what the true
temperature history was; age and temperature trade off against each other
in the forward model (warmer-and-younger vs. colder-and-older can produce
similar D/L). The practical effect on the age posterior is a modestly wider
upper tail rather than a shifted median: folding in temperature uncertainty
is cheap insurance against overconfident intervals, even though it doesn't
change the central estimate much. This supports treating temperature as a
nuisance parameter to marginalize over, as done here, rather than pursuing
tight temperature reconstructions this dataset cannot actually deliver.

## Discussion

All three methods broadly agree: medians are within a few tens of percent
of each other for four of the five validation cases, and intervals overlap
substantially throughout. Darrell's GLS calculator and the empirical
Bayesian model track each other especially closely, which is expected --
the Bayesian version was built to reproduce the same TDK/SPK curves and
combine them coherently rather than to change the underlying calibration.
The forward model, built on a genuinely different (shared, mechanistic)
kinetic assumption, is the one most likely to diverge, and does so most for
Example 1 and the oldest known-age sample, tracing to real modeling choices
rather than a bug:

- **Example 1** has no recorded temperature, so the forward model falls
  back to the dataset mean (`r round(mean_temp, 2)`&deg;C) -- a different
  assumption than the empirical curve fit, which doesn't need a temperature
  input at all.
- **The oldest known-age sample** sits at the edge of the calibration
  range, where both the empirical curve's extrapolation and the forward
  model's `Ea`-driven rate law are least constrained (the original study
  reports a similar systematic bias for its oldest samples via
  leave-one-out cross-validation).

The forward model's real advantage isn't a closer numerical match to the
empirical fit -- it's a **mechanistic** one: `Ea` values around 31 kcal/mol
for both acids are physically interpretable and directly comparable to the
isoleucine kinetics fit elsewhere in this package, and the whole pipeline
(age-depth model + temperature history + Arrhenius kinetics) is ready to
absorb real per-core paleotemperature reconstructions the moment they're
available, rather than needing a new curve refit. Two placeholders stand
between this and a production-ready model: the simulated glacial-
interglacial temperature history (amplitude and period are illustrative,
not fitted or reconstructed), and the straight-line age-depth model used
for the many cores with only one dated sample. Both are natural targets for
follow-up once better site-specific age and temperature constraints are
available -- and the `Ea` compensation-ridge diagnostic above is a direct,
quantitative argument for why heating-experiment data would sharpen this model the most.

The two supplementary analyses above point the same direction: a second
acid helps, but only modestly, because Asp and Glu are highly correlated;
and the temperature history matters for the *width* of the age interval
(via the nuisance-parameter marginalization) more than for its centre, and
this dataset cannot pin it down precisely regardless. Both are reasons to
treat the reported intervals as reasonably conservative rather than
optimistic, and both would benefit from the same real-world inputs called
out above -- multi-temperature heating experiments to separate `Ae` from
`Ea`, and site-specific paleotemperature reconstructions to replace the
simulated placeholder history.
